---
title: "Adding a model to pprof"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a model to pprof}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, dpi = 96, fig.retina = 1, out.width = "100%"
)
```

pprof is built so that a new model plugs into its provider profiling (tests and flags,
standardized measures, intervals, funnel limits, and plots) without any change to that
code. A model supplies its estimates and a few methods, the *model contract*, and declares
which inference it supports. This vignette is for statisticians and programmers who want
to add a model, in pprof itself or in a package of their own. It explains the contract,
builds a small model from scratch, and walks through the design of a larger one.

## How pprof is organized

pprof has five layers, each using only the layers below it:

| Layer | What it does | Functions |
|---|---|---|
| data | parses the formula, drops incomplete rows, builds the design matrix, indexes and screens the providers | `data_prepare()` |
| model | estimates a model and builds its object | `fit_logistic_fe()` and the other fits, `new_pprof_model()` |
| inference | p-values, flags, exact and score tests, test inversion, coefficient tests | used by the profiling functions |
| profiling | tests, measures, effects, funnel limits | `test_providers()`, `standardize_providers()`, `provider_effects()`, `funnel_limits()`, `profile_providers()` |
| presentation | printing, tidy outputs, plots | `print()`, `tidy()`, `plot_funnel()`, ... |

The profiling functions never read a model's fields. They call the contract's generic
functions, and they take what differs between families of models, such as the measures, the
norm, or the funnel limits, from a *family specification* that the model returns. A model
also lists its *capabilities*; asking it for anything else raises an error of class
`pprof_error_unsupported_inference`.

## What a fit function does

1. Check its arguments.
2. Prepare the data with `data_prepare()`: the response, the design matrix, each
   observation's provider, the provider table, and, if asked, the screening of small
   providers.
3. Estimate the model.
4. Build the object with `new_pprof_model()`, which fills the fields every model shares and
   checks them, and adds the model's own fields.
5. Declare its capabilities with a method for `inference_capabilities()`.
6. Implement the contract's generics that have no shared method, and any whose shared
   method does not fit.

## A model in a few lines

The smallest useful model has a binary outcome, no covariates, and an effect for each
provider equal to the logit of its event rate. Its fit function uses `data_prepare()` and
`new_pprof_model()`:

```{r toy-fit, message = FALSE}
library(pprof)

toy_fit <- function(formula, data, provider, min_provider_size = 1) {
  prepared <- data_prepare(formula, data, provider, min_provider_size = min_provider_size, event_counts = TRUE)
  if (ncol(prepared$design) > 0L) stop("The toy model has no covariates.")
  providers <- prepared$providers[prepared$providers$included, , drop = FALSE]
  effects <- stats::qlogis(providers$n_events / providers$n_obs)
  names(effects) <- providers$provider_id
  new_pprof_model(
    prepared,
    coefficients = numeric(),
    vcov = matrix(numeric(), 0L, 0L),
    provider_effects = effects,
    linear_predictor = rep(0, length(prepared$response)),
    spec = list(family = "toy", min_provider_size = min_provider_size),
    class = "pprof_toy_model"
  )
}
```

The model then needs four methods: the inference it supports, the expected outcome of each
observation for a given provider effect, the numeric value of the norm, and its family
specification. A package registers them in its NAMESPACE (see "Registering the methods"
below); a script, like this vignette, registers them with `.S3method()`.

```{r toy-methods}
# The position of each observation's provider among the included providers.
toy_effect_position <- function(model) {
  match(provider_index(model), which(provider_table(model)$included))
}

.S3method("inference_capabilities", "pprof_toy_model", function(model) {
  c("provider_exact", "provider_score", "standardize_indirect", "funnel")
})

.S3method("expected_outcome", "pprof_toy_model", function(model, effect, ...) {
  effect <- if (length(effect) == 1L) rep(effect, length(provider_estimates(model))) else effect
  stats::plogis(effect[toy_effect_position(model)] + linear_predictor(model))
})

.S3method("null_effect", "pprof_toy_model", function(model, null, ...) {
  if (identical(null, "median")) stats::median(provider_estimates(model)) else null
})

.S3method("profile_spec", "pprof_toy_model", function(model) {
  list(
    family = "toy", effect = "gamma", null_default = "median", null_options = "median",
    indirect_numerator = "observed", measures = "ratio",
    variance_function = function(mean) mean * (1 - mean),
    funnel = list(
      measure = "ratio", target = 1, floor = 0, test = "score",
      precision = function(expected, variance) expected^2 / variance,
      half_width = function(critical, precision) critical / sqrt(precision)
    )
  )
})
```

Five hospitals of 10 to 14 patients:

```{r toy-data}
sizes <- 10:14
events <- c(3L, 5L, 2L, 8L, 6L)
toy_data <- data.frame(
  y = unlist(Map(function(n, k) c(rep(1, k), rep(0, n - k)), sizes, events)),
  hospital = rep(sprintf("hospital %d", 1:5), sizes)
)
toy <- toy_fit(y ~ 1, toy_data, "hospital")
toy
```

The profiling functions and the plots now work on it, unchanged:

```{r toy-profiling}
test_providers(toy)$table
measures <- standardize_providers(toy)
measures$table
# The indirect ratio is the number of events over the number expected at the median effect.
all.equal(measures$table$estimate, events / (sizes * plogis(median(provider_estimates(toy)))))
profile <- profile_providers(toy)
plot_funnel(profile)
```

Inference the model does not declare is refused with a classed error:

```{r toy-unsupported}
tryCatch(test_providers(toy, test = "wald"),
         pprof_error_unsupported_inference = function(e) conditionMessage(e))
```

## The model contract

| Generic | Returns | Shared method |
|---|---|---|
| `provider_table(model)` | the provider table: `provider_id`, `provider_value`, `n_obs`, `n_events` (binary outcomes), `included`, `no_events`, `all_events` | yes |
| `provider_estimates(model)` | the effects of the included providers, named by provider ID, in provider order | yes |
| `provider_index(model)` | for each observation, its provider's row in the provider table | yes |
| `linear_predictor(model)` | for each observation, the covariates' linear predictor | yes |
| `observed_outcome(model)` | for each observation, the outcome | yes |
| `inference_capabilities(model)` | the names of the capabilities the model supports | yes, declaring none |
| `expected_outcome(model, effect)` | for each observation, the expected outcome if its provider's effect were `effect` (one value, or one per included provider) | no |
| `null_effect(model, null)` | the number that a choice of `null` (such as `"median"`) stands for | no |
| `profile_spec(model)` | the family specification | no |
| `predicted_outcome(model)` | for each observation, the model's prediction with its own provider effect | no |
| `provider_estimate_se(model)` | the standard errors of the effects | no |
| `provider_test(model, test, ...)` | the per-provider results of a test that needs the model's internals | no |
| `refit_without(model, covariates, data)` | the model refit without some covariates | no |

The shared methods read the fields that `new_pprof_model()` fills. A generic without a
shared method raises `pprof_error_unsupported_inference` for a model that does not
implement it, so a model implements only what its capabilities need.

## The family specification

`profile_spec()` returns a list. Six fields are required:

| Field | Meaning | Logistic FE | Linear FE | Random effects |
|---|---|---|---|---|
| `family` | a name | `"logistic_fe"` | `"linear_fe"` | the model's |
| `effect` | the name of the effect | `"gamma"` | `"gamma"` | `"alpha"` |
| `null_default` | the default norm | `"median"` | `"median"` | `0` |
| `null_options` | the named norms `null_effect()` understands | `"median"` | `"median"`, `"mean"` | none |
| `indirect_numerator` | the sum compared with the expected one: `"observed"` outcomes, or `"predicted"` (then `predicted_outcome()` gives it) | `"observed"` | `"observed"` | `"predicted"` |
| `measures` | the measures | `"ratio"`, `"rate"` | `"difference"` | as the outcome's |

The other fields are needed only by some functions, and those with a default take the
behavior of logistic fixed-effect models when they are left out:

| Field | Needed for | Default |
|---|---|---|
| `variance_function(mean)` | the null variance of indirect standardization, and the funnel precision | none (the variance is then missing) |
| `mean_function(eta)` | intervals of the standardized measures | none |
| `direct_expected(effects, linear_predictor, threads)` | direct standardization | none |
| `funnel` | funnel limits: a list with `measure`, `target`, `floor`, `test`, `precision(expected, variance)` (or with a third argument `n_obs`), and `half_width(critical, precision)` | none |
| `test_default` | the test of `test_providers()` when `test` is not given | `"exact"` |
| `comparison` | `"ratio"` (observed over expected) or `"difference"` (per observation) | `"ratio"` |
| `direct_reference` | the total of direct standardization: `"observed"` or `"null_expected"` | `"observed"` |
| `direct_limits` | how direct limits sum over the population: `"mean_function"` or `"direct_expected"` | `"mean_function"` |
| `one_sided_extremes` | one-sided measure limits for providers with no events or only events | `TRUE` |
| `wald` | the distributions of the Wald tests and intervals: `test` and `interval`, `"normal"` or `"t"`, and `df` | normal |
| `coefficient_wald` | the rule of the coefficients' Wald tests and intervals | normal, two-sided |
| `wald_caution` | warn when Wald inference covers providers with no events or only events | `FALSE` |

`profile_spec()` of a fitted model shows a complete example, for instance
`names(profile_spec(fit_logistic_fe(...)))`.

## Capabilities

| Capability | What it enables | What the model provides |
|---|---|---|
| `provider_exact`, `provider_bootstrap` | the exact and bootstrap tests, for binary outcomes | `expected_outcome()`, `null_effect()` |
| `provider_score` | the modified score test, for binary outcomes | the same |
| `provider_score_standard` | the standard score test | `provider_test(model, "score_standard", ...)` |
| `provider_wald` | Wald tests | `provider_estimate_se()`; `wald` in the specification for t |
| `interval_exact`, `interval_score` | exact and score intervals of the effects and measures, for binary outcomes | `expected_outcome()`; `mean_function` for measures |
| `interval_wald` | Wald intervals | `provider_estimate_se()`; `mean_function` for measures |
| `standardize_indirect` | indirect standardization | `expected_outcome()`; `variance_function` for the variance |
| `standardize_direct` | direct standardization | `direct_expected` |
| `funnel` | funnel limits and `plot_funnel()` | `funnel`, and the test it names |
| `coef_wald` | `confint()`, `summary()`, and Wald coefficient tests | `coefficients` and `vcov` in the object; `coefficient_wald` if not normal |
| `coef_lr`, `coef_score` | likelihood-ratio and score coefficient tests | `refit_without()` |

Declare only what the model supports and what its authors have checked; the capability
list is where that decision is recorded.

## Registering the methods

- **In a script**, as above, with `.S3method(generic, class, method)`.
- **In another package**, with roxygen2's `@exportS3Method` tag on each method, which writes
  the registration into the package's NAMESPACE:

  ```r
  #' @exportS3Method pprof::expected_outcome
  expected_outcome.mymodel <- function(model, effect, ...) {
    # ...
  }
  ```

  The package imports pprof, uses `data_prepare()` and `new_pprof_model()`, and names its
  classes with its own prefix.
- **In pprof itself**, in a file `R/model-<name>.R` with the fit function, the methods, and
  their roxygen tags, following the naming convention of the repository's `dev/NAMING.md`
  (classes `pprof_<name>`, arguments from its vocabulary). No other R file changes.

## Testing a model

- Check that the contract works: the fit validates (`validate_pprof_model()`), the profiling
  functions run, and undeclared inference raises `pprof_error_unsupported_inference`, as
  pprof's own test of this toy model does (`tests/testthat/test-extension-toy-model.R` in the
  repository).
- Check the numbers against an independent implementation where one exists, as pprof checks
  its models against `glm()`, `lm()`, lme4, and logistf.
- Check the edge cases: providers with no events or only events, a single provider, missing
  values, small providers.
- In pprof itself, `R CMD check`, `lintr`, and the architecture test (which checks that each
  file calls only its own and lower layers) must pass.

## Compiled code

An estimation engine in C++ lives in `src/<module>/`, free of Rcpp types, with a thin
adapter `src/rcpp_<module>.cpp` whose exported functions are named
`cpp_<module>_<action>()`. R compiles files in subdirectories only when they are listed, so
each new object file goes into `OBJECTS` in both `src/Makevars` and `src/Makevars.win`; after
changing the adapter, `Rcpp::compileAttributes()` regenerates the glue. Thread counts come
only from a `threads` argument, and parallel regions call no R API.

## Walk-through: a group lasso model

A larger example, as a design rather than code: logistic provider fixed effects with a
group lasso penalty on the covariate coefficients, minimizing
$-\ell(\gamma, \beta) + \lambda \sum_g w_g \lVert \beta_g \rVert_2$ with the provider effects
unpenalized. Nothing in it changes the existing models or the profiling code.

1. **Interface.** `fit_logistic_grouplasso(formula, data, provider, groups = NULL,
   lambda = NULL, n_lambda = 100, lambda_min_ratio = 1e-3, group_weights = NULL,
   standardize = TRUE, min_provider_size = 10, effect_bound = 10, tol = 1e-7,
   max_iter = 10000, threads = 1, verbose = FALSE)`, with the shared arguments named as in
   the other fits.
2. **Data.** `data_prepare()` as it is. The default groups come from the design matrix's
   `assign` attribute, which maps each column to its formula term, so the columns of a factor
   form one group:

   ```{r assign}
   data(ExampleDataBinary)
   example <- data.frame(y = ExampleDataBinary$Y, hospital = ExampleDataBinary$ProvID, ExampleDataBinary$Z)
   example$z5_level <- cut(example$z5, quantile(example$z5, 0:3 / 3), labels = c("low", "mid", "high"),
                           include.lowest = TRUE)
   design <- data_prepare(y ~ z1 + z5_level + I(z1^2), example, "hospital")$design
   colnames(design)
   attr(design, "assign")
   ```

3. **Standardization of the covariates** inside the model's own code, because it belongs to
   this estimator: each group centered and orthonormalized, the coefficients reported on the
   original scale.
4. **Engine.** Block coordinate descent in `src/penalized/`, alternating Newton steps for
   the provider effects (keeping them within `effect_bound` of their median, as the other
   logistic fits do) with proximal steps for each group of coefficients, along a path of
   λ from the smallest value at which every coefficient is 0, with warm starts; an adapter
   `src/rcpp_penalized.cpp`.
5. **Object.** Built with `new_pprof_model()`, with class
   `c("pprof_logistic_grouplasso", "pprof_model")` and its own fields: the λ path, the
   coefficient and effect paths, the degrees of freedom and deviance per λ, the convergence
   per λ, and the selected λ, none at first.
6. **Coefficient paths.** `coef(fit)` returns the path, and `coef(fit, lambda = x)` the
   solution at a value of the path or refit at another; nothing is interpolated, so every
   reported number is a solution.
7. **Selection.** `select_lambda(fit, criterion = c("cv", "bic", "aic"))` sets the selected
   λ and the linear predictor there.
8. **Profiling at the selected λ.** `provider_estimates()` and `linear_predictor()` read
   the selected solution, `expected_outcome()` is the logistic one, and `profile_spec()` is
   that of logistic fixed effects, so the exact and bootstrap tests, the modified score test,
   the standardized measures, and the exact and score intervals work unchanged, conditional on
   the selected covariates. Before a λ is selected, these methods raise
   `pprof_error_invalid_input`.
9. **Capabilities.** Penalization and selection make the information-based variances
   unsuitable, so the model would not declare `coef_wald`, `coef_lr`, `coef_score`,
   `provider_wald`, `interval_wald`, or `provider_score_standard`, and asking for them raises
   `pprof_error_unsupported_inference`. Whether the conditional tests of step 8 should be
   offered at all is a methodological decision for the model's authors, recorded in the
   capability list.
10. **Files.** `R/model-logistic-grouplasso.R` (the fit, the methods, the capabilities),
    `R/model-grouplasso-path.R` (`coef()`, `select_lambda()`), `R/plot-coefficient-path.R`,
    the C++ engine and adapter, two lines in each Makevars file, the tests, and an entry in
    this guide. No existing R file changes.

## Further reading

The repository's `dev/` directory holds the design documents behind this guide:
`dev/design/ARCHITECTURE.md` (the layers, the model objects, the contract, and the
capabilities), `dev/NAMING.md` (names, arguments, classes, and conditions), and
`dev/DEVELOPER_GUIDE.md` (for contributors to pprof itself). The help pages `?data_prepare`,
`?new_pprof_model`, and `?model_contract` document the developer interface.
