---
title: "Migrating from pprof 1.0.3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from pprof 1.0.3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, dpi = 96, fig.retina = 1, out.width = "100%"
)
# After the first one is shown, chunks with hide_deprecation = TRUE leave out the deprecation
# warnings of the old functions, and only those: every other warning is still shown.
local({
  show_warning <- knitr::knit_hooks$get("warning")
  knitr::knit_hooks$set(warning = function(x, options) {
    if (isTRUE(options$hide_deprecation) && grepl("is deprecated as of pprof 2.0.0", x, fixed = TRUE)) return("")
    show_warning(x, options)
  })
})
```

pprof 2.0.0 brings a new interface: one fitting function per model, all taking a formula,
a data frame, and the name of the provider column, and profiling functions that work the
same way for every model. The functions of pprof 1.0.3 are still there and give the same
results, except where pprof 1.0.3 failed, returned wrong results, or attached results to
the wrong providers (listed below). This vignette maps the old functions to the new ones,
shows each step of an analysis both ways, and lists what changed.

The old functions are deprecated: from pprof 2.0.0 on, each warns the first time it is called
in a session, naming the function that replaces it. They remain at least through the next
minor release, and their removal will be announced in the package's NEWS.

## The functions

| pprof 1.0.3 | New | How to migrate |
|---|---|---|
| `logis_fe()`, `logis_firth()` | `fit_logistic_fe()`, `fit_logistic_firth()` | drop `id()` from the formula and name the provider column in `provider`; column names and vectors become a formula and a data frame |
| `linear_fe()` | `fit_linear_fe()` | as above; `option.gamma.var` becomes `provider_variance` |
| `linear_re()`, `logis_re()` | `fit_linear_re()`, `fit_logistic_re()` | drop `(1 | id)` from the formula and name the provider column in `provider` |
| `linear_cre()`, `logis_cre()` | `fit_linear_cre()`, `fit_logistic_cre()` | write `Y.char`, `wb.char`, and `other.char` as a formula, and name `wb.char` in `within_between` |
| `test()` | `test_providers()` | the test names change (below); `parm` becomes `providers` |
| `SM_output()` | `standardize_providers()` | `stdz` becomes `standardization` |
| `confint(option = "gamma")` | `provider_effects(interval = )` | `test` becomes `interval` |
| `confint(option = "SM")` | `standardize_providers(interval = )` | as above |
| `summary()` | `summary()` and `test_coefficients()` | likelihood-ratio and score tests come from `test_coefficients()`, with the data |
| `plot()` (funnel plot) | `plot_funnel(funnel_limits(fit))` | `alpha` becomes `level = 1 - alpha` |
| `caterpillar_plot()` | `plot_caterpillar()` | takes the result of `standardize_providers()` or `provider_effects()` with intervals |
| `bar_plot()` | `plot_flags()` | takes the result of `test_providers()`; `group_num` becomes `group_count` |
| `data_check()` | `check_data()` | write `Y`, `Z`, and `ProvID` as a formula, a data frame, and the provider column; `check_data()` reports every problem instead of stopping at the first |

Every old function remains, as do the `test()`, `SM_output()`, `confint()`, `summary()`,
and `plot()` methods for the objects the old fitting functions return. New in pprof 2.0.0:
`profile_providers()`, which runs the tests, measures, effects, and funnel limits at once;
`plot_volume()`; and `tidy()`, `glance()`, and `augment()`.

## The arguments

| pprof 1.0.3 | New |
|---|---|
| `id()` or `(1 | id)` in the formula; `ProvID.char`; `ProvID` | `provider`, the name of a column of `data` |
| `Y.char`, `Z.char`; `Y`, `Z` | `formula` and `data` |
| `wb.char`, `other.char` | `within_between`; the other covariates are in the formula |
| `method = "SerBIN"`, `"BAN"` | `method = "serbin"`, `"ban"` |
| `max.iter` | `max_iter` |
| `stop = "or"`, `"all"`, `"beta"`, `"relch"`, `"ratch"` | `stop_rule = "any"`, `"all"`, `"coefficients"`, `"relative_loglik"`, `"relative_gain"` |
| `bound` | `effect_bound` |
| `cutoff` | `min_provider_size` |
| `option.gamma.var` | `provider_variance` |
| `message` (default `TRUE`) | `verbose` (default `FALSE`) |
| `parm` (provider functions) | `providers` |
| `test = "exact.poisbinom"`, `"exact.bootstrap"` | `test = "exact"`, `"bootstrap"` |
| `score_modified = TRUE`, `FALSE` | `score_type = "modified"`, `"standard"` |
| `n` | `n_resamples` |
| `stdz` | `standardization` |
| `confint(test = )` | `interval` |
| `alpha` (funnel plot) | `level`, equal to `1 - alpha` |
| `refline_value` | `reference` |
| `group_num` | `group_count` |
| `line_size` | `line_width` |

## Each step, both ways

The examples use the binary example data. The helper `same()` prints `TRUE` when the old
and the new numbers are identical, and stops otherwise; the comparisons run every time this
vignette is built.

```{r data, message = FALSE}
library(pprof)
data(ExampleDataBinary)
example <- data.frame(y = ExampleDataBinary$Y, hospital = ExampleDataBinary$ProvID, ExampleDataBinary$Z)
same <- function(old, new) {
  stopifnot(identical(as.vector(unname(old)), as.vector(unname(new))))
  TRUE
}
```

### Fitting

```{r fit}
# pprof 1.0.3
old <- logis_fe(y ~ id(hospital) + z1 + z2 + z3 + z4 + z5, data = example, message = FALSE)
# pprof 2.0.0
fit <- fit_logistic_fe(y ~ z1 + z2 + z3 + z4 + z5, data = example, provider = "hospital")
same(old$coefficient$beta, coef(fit))
same(old$coefficient$gamma, provider_estimates(fit))
same(old$variance$beta, vcov(fit))
```

The first call of `logis_fe()` warned that it is deprecated. Each old function warns once per
session, the first time it is called, with the class `pprof_deprecated`;
`suppressWarnings(expr, classes = "pprof_deprecated")`, or a calling handler for that class,
silences the warning without silencing others. The rest of this vignette leaves these
warnings out, and shows any other.

```{r hide-deprecation, include = FALSE}
knitr::opts_chunk$set(hide_deprecation = TRUE)
```

The other input formats of pprof 1.0.3, column names and vectors, become a data frame:

```{r fit-formats}
old_columns <- logis_fe(data = example, Y.char = "y", Z.char = paste0("z", 1:5), ProvID.char = "hospital",
                        message = FALSE)
old_vectors <- logis_fe(Y = ExampleDataBinary$Y, Z = ExampleDataBinary$Z, ProvID = ExampleDataBinary$ProvID,
                        message = FALSE)
same(old_columns$coefficient$beta, coef(fit))
same(old_vectors$coefficient$beta, coef(fit))
```

The new model keeps less than the old object (no copy of the data, unless
`keep_data = TRUE`) and more about the fit: `fit$convergence` records the iterations and
the convergence criteria, and a fit that reaches the iteration limit warns.

### Testing providers

```{r tests}
old_tests <- test(old)
tests <- test_providers(fit)
same(old_tests[["p value"]], tests$table$p_value)
same(as.integer(as.character(old_tests$flag)), tests$table$flag)
head(tests$table, 3)
```

The old result is a data frame with the providers as row names and the flag as a factor
whose levels are the flags that occur; the new result's `table` has a `provider_id` column
and integer flags (-1, 0, 1), and the result records its settings (`tests$test`,
`tests$level`, `tests$null_value`, ...).

### Standardized measures

```{r measures}
old_measures <- SM_output(old)
measures <- standardize_providers(fit)
same(old_measures$indirect.ratio, measures$table$estimate[measures$table$measure == "ratio"])
same(old_measures$indirect.rate, measures$table$estimate[measures$table$measure == "rate"])
```

The new table has one row per standardization, measure, and provider, with the observed
and expected counts that pprof 1.0.3 kept in `OE`.

### Intervals

`confint()` of an old object gives intervals for provider effects or standardized
measures; for the new models, these come from `provider_effects()` and
`standardize_providers()`, and `confint()` gives the intervals of the covariate
coefficients, as for other models in R. Here for the first ten hospitals:

```{r intervals}
old_effects <- confint(old, parm = 1:10, option = "gamma")
effects <- provider_effects(fit, interval = "exact", providers = 1:10)
same(c(old_effects$gamma.lower, old_effects$gamma.upper), c(effects$table$lower, effects$table$upper))
old_ratios <- confint(old, parm = 1:10)$CI.indirect_ratio
ratios <- standardize_providers(fit, measure = "ratio", interval = "exact", providers = 1:10)
same(c(old_ratios$CI_ratio.lower, old_ratios$CI_ratio.upper), c(ratios$table$lower, ratios$table$upper))
```

### Covariates

```{r covariates}
old_summary <- summary(old)
same(c(old_summary$Estimate, old_summary$CI.Lower, old_summary$CI.Upper), c(coef(fit), confint(fit)))
old_lr <- summary(old, test = "lr")
same(old_lr$stat, test_coefficients(fit, test = "lr", data = example)$table$statistic)
```

The new `summary()` returns numbers, which it prints formatted, where the old one gave the
p-values as text. The likelihood-ratio and score tests refit the model without each
covariate; since the new model does not keep the covariates, they need `data` (or a fit
with `keep_data = TRUE`).

### Plots

The old plot functions draw what they drew in pprof 1.0.3. The new ones take the new
results, give each flag the same colour in every plot, and return ggplot objects:

```{r plots}
old_funnel <- plot(old)
funnel <- funnel_limits(fit)
same(sort(old_funnel$layers[[1]]$data$indicator), sort(funnel$providers$estimate))
plot_funnel(funnel)
```

`caterpillar_plot(confint(old)$CI.indirect_ratio)` becomes
`plot_caterpillar(standardize_providers(fit, interval = "exact"))`, and
`bar_plot(test(old))` becomes `plot_flags(test_providers(fit))`.

### Other models

The other fitting functions translate in the same way:

```{r other-models, message = FALSE}
data(ExampleDataLinear)
continuous <- data.frame(y = ExampleDataLinear$Y, hospital = ExampleDataLinear$ProvID, ExampleDataLinear$Z)
risk <- y ~ z1 + z2 + z3 + z4 + z5

old_fe <- linear_fe(y ~ id(hospital) + z1 + z2 + z3 + z4 + z5, data = continuous)
same(old_fe$coefficient$beta, coef(fit_linear_fe(risk, continuous, "hospital")))

old_re <- linear_re(y ~ (1 | hospital) + z1 + z2 + z3 + z4 + z5, data = continuous)
same(old_re$coefficient$RE, provider_estimates(fit_linear_re(risk, continuous, "hospital")))

old_cre <- linear_cre(data = continuous, Y.char = "y", wb.char = c("z1", "z2"), other.char = c("z3", "z4", "z5"),
                      ProvID.char = "hospital")
same(old_cre$coefficient$FE, coef(fit_linear_cre(risk, continuous, "hospital", within_between = c("z1", "z2"))))

old_firth <- logis_firth(y ~ id(hospital) + z1 + z2 + z3 + z4 + z5, data = example, message = FALSE)
same(old_firth$coefficient$beta, coef(fit_logistic_firth(risk, example, "hospital")))
```

`logis_re()` and `logis_cre()` translate like `linear_re()` and `linear_cre()`.

## What is different in the new interface

- **One way to give the data.** A formula, a data frame, and the name of the provider
  column, for every model. Formulas may contain transformations, interactions, and factors.
- **Results.** Every profiling function returns a result with a `table` (a data frame
  keyed by `provider_id`, in provider order) and the settings used; flags are the integers
  -1, 0, and 1. `tidy()` turns a result or a model into a tibble.
- **`confint()`** of a model gives the intervals of the covariate coefficients; intervals
  for providers come from `provider_effects()` and `standardize_providers()`.
- **Quiet by default.** The new fitting functions print nothing unless `verbose = TRUE`;
  screening and the iteration limit still raise warnings. The old functions print as
  they did.
- **One thread by default.** The new functions that take `threads` use one thread unless
  asked for more, where pprof 1.0.3's `SM_output()` used two by default. Results do not
  depend on the number of threads.
- **Compact models.** A model keeps what the profiling needs, but not the matrix of
  covariates; the standard score test and the likelihood-ratio and score tests of the
  coefficients take `data`, or a fit with `keep_data = TRUE`.
- **Conditions with classes.** Errors and warnings have classes (for example
  `pprof_error_unsupported_inference` when a model does not support the inference asked for,
  or `pprof_warning_screening`), so code can catch them by class.
- **Defaults.** `null = NULL` means the model's norm (the median effect for fixed-effect
  models, 0 for random-effect models); providers are selected by ID, compared as text, so
  integer, character, and factor IDs work alike.

## What changed in the old functions

The old functions return the same objects and numbers as pprof 1.0.3, including its
defaults and its conventions, with these exceptions, where pprof 1.0.3 failed, returned a
wrong result, or misaligned results:

- **Fitting.** Formulas with transformed terms, interactions, or factor levels containing
  spaces work. Factor provider IDs work when providers are excluded by screening. Settings
  that returned meaningless fits now raise errors: `threads` below 1, `backtrack` other
  than 0 or 1 with BAN, outcomes other than 0 and 1, and non-positive `max.iter`, `tol`, or
  `bound`. `logis_firth()` with more than one thread gives the result of one thread, where
  it stopped after a few iterations, and a singular information matrix raises an error
  instead of ending the R session. In `linear_re()` and `logis_re()`, character provider IDs
  with a covariate matrix, and calls that match no input format, raise errors instead of
  failing in lme4 or with "object 'fit_re' not found".
- **Messages.** The screening warning counts only the providers excluded, a fit that
  reaches the iteration limit says "not converged" and warns, attaching pprof no longer
  prints a message about a method of the car package, pROC's messages no longer
  appear, and neither do the ggplot2 warnings of `bar_plot()` and `caterpillar_plot()`
  (the horizontal caterpillar plot still gives ggplot2's message about
  `geom_errorbarh()`).
- **Tests and intervals.** With character provider IDs, `confint()` keeps each provider's
  interval with that provider (pprof 1.0.3 attached the intervals of the providers after
  the first one without events to other providers), and it works with factor provider IDs
  and with a provider column of any name. Providers stored as integers can be selected with `parm`, and
  an integer `null` works in every method. A `null` of several values raises an error
  instead of being applied by position. `test(test = "robust_wald")`, which returned
  `NULL`, raises an error. A standard score statistic that cannot be computed gives a missing
  p-value and flag with a warning, where `test()` failed. `summary(test = "lr")` and
  `"score"` work with two covariates (with one, they raise an error), and the summaries of
  random-effect fits work on objects without their lme4 fit. The `confint()` methods of
  `logis_re()` and `logis_cre()` fits use one thread, where they used four.
- **Plots.** `plot(test = "exact")`, which always failed, raises a classed error. Inputs
  that `caterpillar_plot()` and `bar_plot()` failed on raise classed errors, and the plot
  data of `bar_plot()` are a data frame rather than a grouped tibble.
- **Data checks.** `data_check()` completes its checks with a single covariate, where it
  stopped with an error at the correlation check, and its errors have the class
  `pprof_error_data`.

Nothing else changes: estimates, tests, p-values, intervals, flags, standardized measures,
and defaults are those of pprof 1.0.3.
