---
title: "Models and what they support"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models and what they support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, dpi = 96, fig.retina = 1, out.width = "100%"
)
```

pprof fits seven models. Every fit takes a formula with the outcome and the risk
factors, a data frame, and the name of the column that identifies the providers, and
returns a model that the profiling functions (`test_providers()`,
`standardize_providers()`, `provider_effects()`, `funnel_limits()`, and
`profile_providers()`) work with. This vignette describes the models, what each assumes,
and what each supports. `vignette("pprof", package = "pprof")` shows a complete analysis,
and `vignette("statistical-methods", package = "pprof")` gives the formulas.

```{r data, message = FALSE}
library(pprof)
data(ExampleDataBinary)
data(ExampleDataLinear)
binary <- data.frame(y = ExampleDataBinary$Y, hospital = ExampleDataBinary$ProvID, ExampleDataBinary$Z)
continuous <- data.frame(y = ExampleDataLinear$Y, hospital = ExampleDataLinear$ProvID, ExampleDataLinear$Z)
risk <- y ~ z1 + z2 + z3 + z4 + z5
```

## The seven models

| Outcome | Fixed effects | Random effects | Correlated random effects |
|---|---|---|---|
| binary | `fit_logistic_fe()`, `fit_logistic_firth()` | `fit_logistic_re()` | `fit_logistic_cre()` |
| continuous | `fit_linear_fe()` | `fit_linear_re()` | `fit_linear_cre()` |

- **Fixed effects.** Each provider's effect γ~i~ is a parameter, estimated from the
  provider's own observations: by maximum likelihood for `fit_logistic_fe()`, by Firth's
  penalized likelihood for `fit_logistic_firth()`, and by least squares for
  `fit_linear_fe()`. Nothing is assumed about how the effects vary across providers.
- **Random effects.** The effects are α~i~, drawn from a normal distribution with mean 0
  and variance σ^2^~α~ around an overall intercept, and the model is fitted by lme4
  (Bates et al. 2015): `lmer()` for continuous outcomes, by restricted maximum likelihood,
  and `glmer()` for binary outcomes, by the Laplace approximation. The reported effects are
  the conditional modes, which are shrunk toward 0, the more so the fewer observations a
  provider has. The model assumes that the effects are independent of the risk factors.
- **Correlated random effects.** The random-effect model, with each risk factor named in
  `within_between` split into its provider mean and the deviation from that mean, each
  with its own coefficient (Mundlak 1978). The effects may then be correlated with the
  provider means of those risk factors.

### Choosing a model

The models answer different questions, and pprof leaves the choice to the analyst; these
points, from the literature, bear on it.

- If the provider effects are correlated with the risk factors, for example because
  sicker patients go to particular hospitals, the random-effect model's estimates of the
  coefficients, and with them of the effects, are biased, while the fixed-effect model's
  are not, and the correlated random-effect model removes the correlation with the provider
  means of the decomposed risk factors (Mundlak 1978; Hsiao 2022). In a linear model with
  every risk factor decomposed, the within-provider coefficients of the correlated
  random-effect model are the fixed-effect coefficients (shown below).
- The random-effect estimates borrow strength from the other providers and are shrunk
  toward the overall average. Profiling is about the providers at the extremes, and Xia et
  al. (2022) prefer the fixed-effect estimates for profiling when the variation between
  providers is entirely due to the quality of their care.
- The tests and intervals of random-effect models compare the shrunk estimates with their
  conditional standard deviations; their flags answer a different question from the
  fixed-effect flags and are not interchangeable with them.
  `vignette("statistical-methods", package = "pprof")` describes how often they flag a
  provider and how often their intervals cover its effect.
- For binary outcomes, Firth's penalty reduces the small-sample bias of the maximum
  likelihood estimates and keeps the estimates finite when a provider's patients all have
  the same outcome (Firth 1993; Heinze and Schemper 2002), which matters for small
  providers and rare events.

## What each model supports

Each model declares the inference it supports, and the profiling functions refuse, with a
`pprof_error_unsupported_inference` error, what a model does not declare. The table below
is computed from fits of the example data (the logistic random-effect models on the first
20 hospitals, to keep this vignette fast).

```{r fits, message = FALSE}
small <- binary[binary$hospital <= 20, ]
fits <- list(
  fit_logistic_fe = fit_logistic_fe(risk, binary, "hospital"),
  fit_logistic_firth = fit_logistic_firth(risk, binary, "hospital"),
  fit_logistic_re = fit_logistic_re(risk, small, "hospital"),
  fit_logistic_cre = fit_logistic_cre(risk, small, "hospital", within_between = c("z1", "z2")),
  fit_linear_fe = fit_linear_fe(risk, continuous, "hospital"),
  fit_linear_re = fit_linear_re(risk, continuous, "hospital"),
  fit_linear_cre = fit_linear_cre(risk, continuous, "hospital", within_between = c("z1", "z2"))
)
declared <- function(fit, prefix) {
  capabilities <- inference_capabilities(fit)
  paste(sub(prefix, "", grep(prefix, capabilities, value = TRUE)), collapse = ", ")
}
support <- data.frame(
  model = names(fits),
  provider_tests = vapply(fits, declared, "", prefix = "^provider_"),
  default_test = vapply(fits, function(fit) {
    test <- profile_spec(fit)$test_default
    if (is.null(test)) "exact" else test
  }, ""),
  intervals = vapply(fits, declared, "", prefix = "^interval_"),
  default_null = vapply(fits, function(fit) as.character(profile_spec(fit)$null_default), ""),
  measures = vapply(fits, function(fit) paste(profile_spec(fit)$measures, collapse = ", "), ""),
  indirect_numerator = vapply(fits, function(fit) profile_spec(fit)$indirect_numerator, ""),
  funnel = vapply(fits, function(fit) "funnel" %in% inference_capabilities(fit), NA),
  coefficient_tests = vapply(fits, declared, "", prefix = "^coef_"),
  row.names = NULL
)
knitr::kable(support)
```

Three differences between the families are worth stating:

- Indirect measures compare a provider's observed outcomes with those expected under the
  norm for the fixed-effect models, but its predicted outcomes, which include its estimated
  effect, for the random-effect models ("predicted" in the table), as in pprof 1.0.3.
- Logistic models give ratios and rates; linear models give differences.
- The norm is the median of the provider effects for the fixed-effect models (linear
  fixed-effect models also offer `null = "mean"`), and 0 for the random-effect models;
  `null` can always be a number.

## Fixed-effect models

### Small providers and providers with extreme outcomes

`fit_logistic_fe()` and `fit_logistic_firth()` exclude providers with fewer than
`min_provider_size` complete observations (10 by default), with a warning of class
`pprof_warning_screening`; the other models keep every provider. The example's hospitals
have at least 50 patients each, so a larger threshold shows the screening:

```{r screening}
screened <- fit_logistic_fe(risk, binary, "hospital", min_provider_size = 60)
screened$n_excluded_providers
```

A provider without events has an effect whose maximum likelihood estimate is minus
infinity, and one with only events plus infinity. `fit_logistic_fe()` keeps every effect
within `effect_bound` of the median effect, so these effects stay finite, but each
iteration moves them further toward the bound, and where they end depends on when the fit
stops. The example has three hospitals without events:

```{r extreme}
fit <- fits$fit_logistic_fe
providers <- provider_table(fit)
no_events <- providers$provider_id[providers$no_events]
tight <- fit_logistic_fe(risk, binary, "hospital", stop_rule = "all", tol = 1e-10)
data.frame(
  hospital = no_events,
  default_fit = provider_estimates(fit)[no_events],
  tight_fit = provider_estimates(tight)[no_events],
  firth = provider_estimates(fits$fit_logistic_firth)[no_events],
  bound = median(provider_estimates(fit)) - 10,
  row.names = NULL
)
```

The default fit stops after `r fit$convergence$iterations` iterations, before these
effects reach the bound; the tight fit takes `r tight$convergence$iterations` and puts all
three at the bound. Firth's estimates are finite maxima of the penalized likelihood. The
exact and score tests, their intervals, and the indirect measures use these effects only
through the median effect, the norm, which does not change while they stay below it; Wald
tests and intervals use them directly, so they warn when they cover such providers.

### Convergence

The fit stops when its stopping rule's criterion falls below `tol`. There are three
criteria: the largest change of a coefficient, the change of the log-likelihood relative
to its value, and the change relative to the gain since the starting values. The default
rule, `"any"`, stops as soon as the smallest of the three is below `tol`, as pprof 1.0.3
does; `"all"` waits for the largest. The log-likelihood criteria shrink as the data grow,
so on large data a fit with `"any"` can stop while the provider effects still change. The
object keeps the criteria of every iteration:

```{r convergence}
fit$convergence$criteria
max(abs(coef(tight) - coef(fit)))
```

On large data, check these criteria, or compare the fit with one using
`stop_rule = "all"`.

### SerBIN and BAN

`method = "serbin"` (the default) updates the provider effects and the coefficients
together, inverting the information matrix through its block structure (Wu et al. 2022);
`method = "ban"` alternates between them (He et al. 2013). Both maximize the same
likelihood and differ only in the path they take and where their stopping rule stops them:

```{r methods}
ban <- fit_logistic_fe(risk, binary, "hospital", method = "ban")
c(serbin = fit$convergence$iterations, ban = ban$convergence$iterations)
max(abs(coef(ban) - coef(fit)))
```

### Firth's penalty

`fit_logistic_firth()` maximizes the penalized likelihood of Firth (1993). Its models
work with every function for `fit_logistic_fe()` models. As in pprof 1.0.3, the variances,
the log-likelihood, AIC, and BIC are those of the unpenalized likelihood at the Firth
estimates; the penalized log-likelihood is kept as `penalized_loglik`.

### Linear fixed effects

`fit_linear_fe()` estimates the coefficients by least squares within providers and each
provider's effect from its mean outcome. `provider_variance` chooses the variance of the
estimated effects: `"simplified"` (the default) σ^2^ / n~i~, or `"full"`, which adds the
uncertainty of the coefficients. The Wald tests use the normal distribution with the
simplified variance and Student's t with the full variance; the intervals do the reverse,
as in pprof 1.0.3.

```{r linear-fe}
full <- fit_linear_fe(risk, continuous, "hospital", provider_variance = "full")
head(data.frame(
  simplified = provider_effects(fits$fit_linear_fe)$table$std_error,
  full = provider_effects(full)$table$std_error
))
```

## Random-effect models

`fit_linear_re()` and `fit_logistic_re()` call lme4 with the formula
`outcome ~ risk factors + (1 | provider)`, and pass any further arguments, such as
`control`, to `lmer()` or `glmer()`. The object keeps lme4's convergence messages in
`fit$convergence`, and the lme4 fit itself with `keep_data = TRUE`.

When lme4 estimates the provider variance as 0 (a singular fit), every provider effect and
its standard deviation are 0, the test statistics are undefined, and every flag is
missing, as in pprof 1.0.3. `plot_flags()` shows such providers as "no flag". Simulated
data without provider effects show it:

```{r singular, message = FALSE}
set.seed(1)
simulated <- data.frame(clinic = rep(1:20, each = 10), x = rnorm(200))
simulated$y <- 1 + 0.5 * simulated$x + rnorm(200)
singular <- fit_linear_re(y ~ x, simulated, "clinic")
singular$convergence$singular
table(test_providers(singular)$table$flag, useNA = "ifany")
```

## Correlated random-effect models

`fit_linear_cre()` and `fit_logistic_cre()` replace each risk factor named in
`within_between` by its provider mean (`<name>_bar`) and its deviation from that mean
(`<name>_within`). As in pprof 1.0.3, the provider means are computed over every row of the
data, before rows with missing values are dropped.

With every risk factor decomposed, the within-provider coefficients of the linear model
are the fixed-effect coefficients:

```{r cre}
decomposed <- fit_linear_cre(risk, continuous, "hospital", within_between = paste0("z", 1:5))
within <- coef(decomposed)[paste0("z", 1:5, "_within")]
max(abs(within - coef(fits$fit_linear_fe)))
```

## Methods that need the risk factors

Models keep, for every observation, what the profiling needs (the outcome, the linear
predictor, the provider), but not the matrix of risk factors. The standard score test and
the likelihood-ratio and score tests of the coefficients need it: fit with
`keep_data = TRUE`, or pass the data again.

```{r data-required}
standard <- test_providers(fit, test = "score", score_type = "standard", data = binary)
head(standard$table, 3)
```

Without either, these functions stop with an error of class `pprof_error_data_required`.

## Other methods

Models have the usual methods: `coef()`, `vcov()`, `confint()`, `summary()`, `nobs()`,
`formula()`, `logLik()` (and so `AIC()` and `BIC()`), `fitted()`, `residuals()`,
`predict()` (also for new data), and `tidy()`, `glance()`, and `augment()`.

```{r methods-of-models}
predict(fit, newdata = binary[c(1, 500, 1000), ], type = "response")
AIC(fit)
```

## References

Bates D, Mächler M, Bolker B, Walker S (2015). Fitting linear mixed-effects models using
lme4. *Journal of Statistical Software*, 67(1), 1-48. <https://doi.org/10.18637/jss.v067.i01>

Firth D (1993). Bias reduction of maximum likelihood estimates. *Biometrika*, 80(1),
27-38. <https://doi.org/10.1093/biomet/80.1.27>

He K, Kalbfleisch JD, Li Y, Li Y (2013). Evaluating hospital readmission rates in dialysis
facilities; adjusting for hospital effects. *Lifetime Data Analysis*, 19, 490-512.
<https://doi.org/10.1007/s10985-013-9264-6>

Heinze G, Schemper M (2002). A solution to the problem of separation in logistic
regression. *Statistics in Medicine*, 21(16), 2409-2419. <https://doi.org/10.1002/sim.1047>

Hsiao C (2022). *Analysis of Panel Data*, 4th edition. Cambridge University Press.

Mundlak Y (1978). On the pooling of time series and cross section data. *Econometrica*,
46(1), 69-85. <https://doi.org/10.2307/1913646>

Wu W, Yang Y, Kang J, He K (2022). Improving large-scale estimation and inference for
profiling health care providers. *Statistics in Medicine*, 41(15), 2840-2853.
<https://doi.org/10.1002/sim.9387>

Xia L, He K, Li Y, Kalbfleisch JD (2022). Accounting for total variation and robustness in
profiling health care providers. *Biostatistics*, 23(1), 257-273.
<https://doi.org/10.1093/biostatistics/kxaa024>
