---
title: "Getting started with pprof"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pprof}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, dpi = 96, fig.retina = 1, out.width = "100%"
)
```

Provider profiling compares health care providers, such as hospitals, dialysis
facilities, or transplant centers, on an outcome of the people they treat, after
adjusting for those people's characteristics, which the providers do not control. The
same methods apply to other units, such as schools. pprof fits a risk-adjustment model
with an effect for each provider, and then, for every provider, tests whether its outcomes
are higher or lower than expected, computes standardized measures with intervals, and
draws plots of the results.

This vignette goes through an analysis of the simulated binary outcomes in
`ExampleDataBinary`, then of a continuous outcome. The other vignettes go further:
`vignette("models", package = "pprof")` describes the seven models and what each
supports, `vignette("statistical-methods", package = "pprof")` gives the formulas behind
every result, and `vignette("migration", package = "pprof")` shows how code written for
pprof 1.0.3 maps to the functions used here.

## The data

pprof works on a data frame with one row per observation (here, a patient): the outcome,
the risk factors, and a column that identifies the provider. The bundled example is a
list, so we put it into a data frame first.

```{r data, message = FALSE}
library(pprof)
data(ExampleDataBinary)
example <- data.frame(y = ExampleDataBinary$Y, hospital = ExampleDataBinary$ProvID, ExampleDataBinary$Z)
head(example)
```

```{r data-sizes, include = FALSE}
sizes <- table(example$hospital)
```

`example` holds `r nrow(example)` patients of `r length(sizes)` hospitals, with
`r min(sizes)` to `r max(sizes)` patients each, a binary outcome `y`, and five risk
factors `z1` to `z5`.

## Fitting a model

For a binary outcome, `fit_logistic_fe()` fits the logistic model
logit P(y~ij~ = 1) = γ~i~ + z~ij~'β, with a fixed effect γ~i~ for each hospital i and
coefficients β for the risk factors, by maximum likelihood, with the serial blockwise
inversion Newton algorithm (SerBIN, Wu et al. 2022) or the block ascent Newton algorithm
(He et al. 2013). The formula gives the outcome and the risk factors; the provider column
is named separately.

```{r fit}
fit <- fit_logistic_fe(y ~ z1 + z2 + z3 + z4 + z5, data = example, provider = "hospital")
fit
```

```{r fit-facts, include = FALSE}
providers <- provider_table(fit)
```

Before fitting, `fit_logistic_fe()` drops rows with a missing value, and it excludes
providers with fewer than `min_provider_size` complete observations (10 by default), with
a warning; no hospital of the example is excluded. Providers whose patients all have the
same outcome are kept, although their effects cannot be estimated: the maximum likelihood
estimate of the effect of a provider without events is minus infinity. pprof keeps every
effect within `effect_bound` (10 by default) of the median effect, so these estimates stay
finite, but they depend on when the iterations stop, and Wald tests and intervals are
unreliable for them. The example has `r sum(providers$no_events)` hospitals without
events and `r sum(providers$all_events)` with only events.

The fit stops when its convergence criterion falls below `tol`, and `fit$convergence`
records the iterations and the criteria. The default stopping rule, `"any"`, stops as soon
as one of three criteria is below `tol`; on large data, check the criteria in
`fit$convergence`, or fit again with `stop_rule = "all"`, which waits for all three
(see `vignette("models", package = "pprof")`).

## Covariates

`summary()` and `confint()` give Wald tests and intervals for the coefficients of the
risk factors, as `summary()` did in pprof 1.0.3:

```{r covariates}
fit_summary <- summary(fit)
fit_summary
confint(fit)
```

`test_coefficients()` also gives likelihood-ratio and score tests. They refit the model
without each covariate, so they need the data:

```{r covariates-lr}
test_coefficients(fit, test = "lr", data = example)$table
```

## Testing providers

`test_providers()` tests each hospital's effect against a norm, by default the median of
the hospital effects, and flags the hospital as higher than expected (1), as expected (0),
or lower than expected (-1). For logistic fixed-effect models, the default test is exact:
under the null hypothesis, a hospital's number of events is a sum of independent
Bernoulli outcomes, one per patient, with the probabilities the model gives its patients
at the norm, so it has a Poisson-binomial distribution.

```{r tests}
tests <- test_providers(fit)
head(tests$table)
table(tests$table$flag)
```

At level 0.95, `r sum(tests$table$flag == 1)` hospitals are flagged as higher and
`r sum(tests$table$flag == -1)` as lower than expected. `test = "score"`, `"wald"`, and
`"bootstrap"` choose other tests, and `level`, `alternative`, `null`, and `providers` the
level, a one-sided alternative, another norm, and the hospitals to report.

Each hospital is tested on its own, at the chosen level, without an adjustment for
multiple testing. Even if no hospital differed from the norm, about 1 in 20 hospitals
would be flagged at level 0.95, here about `r round(0.05 * length(sizes))` of
`r length(sizes)`. Jones, Ohlssen and Spiegelhalter (2008) discuss false discovery rates
for comparing many providers.

## Standardized measures

A standardized measure compares a hospital's outcomes with the outcomes expected for its
patients. By default, `standardize_providers()` computes indirectly standardized ratios
and rates:

```{r measures}
measures <- standardize_providers(fit)
head(measures$table)
```

For hospital i, the observed count O~i~ is its number of events, and the expected count
E~i~ is the number of events its patients would have if the hospital's effect were the
norm, the median effect: the number expected if hospital i performed at the norm. The
indirectly standardized ratio is O~i~ / E~i~; a ratio above 1 means more events than
expected for these patients. The rate is the ratio times the overall event rate,
`r round(measures$population_rate, 2)`%, kept within 0% and 100%.

Direct standardization, `standardization = "direct"`, applies each hospital's effect to
all the patients in the data: the ratio is the number of events the whole population
would have with that hospital's effect, over the number it has. The two answer different
questions. An indirect ratio compares a hospital with the norm for its own patients, so
two hospitals with different patients are compared with different expectations; direct
ratios apply every hospital's effect to the same patients, so they put the hospitals on a
common footing.

With `interval`, the table also has the limits of an interval. The exact and score
intervals come from inverting the corresponding tests:

```{r measures-intervals}
measures <- standardize_providers(fit, interval = "exact")
head(measures$table[measures$table$measure == "ratio", ])
```

## Provider effects

`provider_effects()` gives the estimated effects themselves, with their standard errors
and, optionally, intervals, here for the first six hospitals:

```{r effects}
effects <- provider_effects(fit, interval = "exact", providers = 1:6)
effects$table
```

## Everything at once

`profile_providers()` runs the tests, the measures, the effects, and, where the model
supports it, the funnel limits, and gathers what a report needs in one table:

```{r profile}
profile <- profile_providers(fit)
head(profile$table)
```

The component results are kept as `profile$tests`, `profile$measures`, `profile$effects`,
and `profile$funnel`; `interval` adds intervals to the measures.

## Plots

Every plot function takes a result and returns a ggplot object, which ggplot2's functions
can change further.

A funnel plot shows each hospital's indirect ratio against its precision, with control
limits around 1 (one or more levels: `funnel_limits(fit, level = c(0.95, 0.998))`):

```{r funnel}
plot_funnel(profile)
```

```{r funnel-facts, include = FALSE}
score_flags <- test_providers(fit, test = "score")$table$flag
```

The limits are those of the score test, and the colours are its flags at the first level,
so a hospital is flagged when its point lies outside those limits (a test-specific funnel
plot, Wu et al. 2023). The score test's flags can differ from the exact test's, here for
`r sum(score_flags != tests$table$flag)` hospital.

A caterpillar plot shows the measures with their intervals, in increasing order, against
the reference line:

```{r caterpillar}
plot_caterpillar(measures, measure = "ratio", use_flag = TRUE)
```

`plot_flags()` shows the shares of flags in groups of hospitals of similar size, and
`plot_volume()` the measures against the number of patients:

```{r flags-volume}
plot_flags(tests)
plot_volume(profile)
```

## Tidy outputs

`tidy()`, `glance()`, and `augment()` turn models and results into tibbles:

```{r tidy}
tidy(fit)
glance(fit)
head(augment(fit))
```

## A continuous outcome

`ecls_data` holds the math scores of fifth-grade children and the schools they attend.
`fit_linear_fe()` fits a linear model with a fixed effect for each school:

```{r ecls}
data(ecls_data)
ecls_fit <- fit_linear_fe(Math_Score ~ Income + Child_Sex, data = ecls_data, provider = "School_ID")
coef(ecls_fit)
```

```{r ecls-sizes, include = FALSE}
school_sizes <- table(ecls_data$School_ID)
```

For a continuous outcome, the measures are differences: a school's indirect difference is
its mean observed score minus its mean expected score. Of the `r length(school_sizes)`
schools, `r sum(school_sizes == 1)` have a single child, whose effect rests on one score.
As the vignette of pprof 1.0.3 did, the plots below show the
`r sum(school_sizes >= 5)` schools with at least five children.

```{r ecls-profile}
larger <- names(school_sizes)[school_sizes >= 5]
ecls_profile <- profile_providers(ecls_fit, interval = "wald", providers = larger)
plot_funnel(ecls_profile)
plot_flags(ecls_profile$tests)
```

## References

He K, Kalbfleisch JD, Li Y, Li Y (2013). Evaluating hospital readmission rates in dialysis
facilities; adjusting for hospital effects. *Lifetime Data Analysis*, 19, 490-512.
<https://doi.org/10.1007/s10985-013-9264-6>

Jones HE, Ohlssen DI, Spiegelhalter DJ (2008). Use of the false discovery rate when
comparing multiple health care providers. *Journal of Clinical Epidemiology*, 61(3),
232-240. <https://doi.org/10.1016/j.jclinepi.2007.04.017>

Wu W, Kuriakose JP, Weng W, Burney RE, He K (2023). Test-specific funnel plots for
healthcare provider profiling leveraging individual- and summary-level information.
*Health Services and Outcomes Research Methodology*, 23(1), 45-58.
<https://doi.org/10.1007/s10742-022-00285-9>

Wu W, Yang Y, Kang J, He K (2022). Improving large-scale estimation and inference for
profiling health care providers. *Statistics in Medicine*, 41(15), 2840-2853.
<https://doi.org/10.1002/sim.9387>
