---
title: "Statistical methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette states what pprof computes: the models, how they are fitted, and the
formulas of its tests, standardized measures, intervals, and funnel limits, with the
numerical conventions of the code where they matter. The results are those of pprof 1.0.3,
which pprof reproduces. A few short computations check formulas against the package.
`vignette("pprof", package = "pprof")` shows the functions in use.

```{r data, message = FALSE}
library(pprof)
data(ExampleDataBinary)
example <- data.frame(y = ExampleDataBinary$Y, hospital = ExampleDataBinary$ProvID, ExampleDataBinary$Z)
fit <- fit_logistic_fe(y ~ z1 + z2 + z3 + z4 + z5, data = example, provider = "hospital")
```

## Notation

Provider $i = 1, \ldots, m$ has observations $j = 1, \ldots, n_i$, and $n = \sum_i n_i$.
Observation $ij$ has the outcome $Y_{ij}$ and the covariates (risk factors) $Z_{ij}$, a
vector of length $p$ with coefficients $\beta$. Provider effects are $\gamma_i$ in
fixed-effect models and $\alpha_i$ in random-effect models. Tests and intervals use
$a = 1 - {}$`level` (computed in floating point, so $1 - 0.95$ is not exactly $0.05$),
$z_q$ is the $q$ quantile of the standard normal distribution, $\Phi$ its distribution
function, and $\operatorname{plogis}(x) = 1 / (1 + e^{-x})$.

## Models

Risk adjustment accounts for the patient characteristics that providers do not control,
such as age and comorbidities, so that providers are compared on the outcomes expected for
their own patients rather than on raw averages. It accounts only for the characteristics
in the model: differences between providers' patients that the covariates do not capture
remain in the provider effects, and may explain much of the variation between providers
(Xia et al. 2022).

**Fixed effects.** For binary and continuous outcomes,
$$
\operatorname{logit} P(Y_{ij} = 1) = \gamma_i + Z_{ij}^\top \beta,
\qquad
Y_{ij} = \gamma_i + Z_{ij}^\top \beta + \epsilon_{ij}, \quad \epsilon_{ij} \sim N(0, \sigma^2).
$$
An overall intercept $\mu$ and provider deviations $\alpha_i$ are not separately
identifiable in these models, so the effects $\gamma_i = \mu + \alpha_i$ are estimated
directly, and the design matrix has no intercept.

**Random effects.** The effects are $\alpha_i \sim N(0, \sigma_\alpha^2)$, independent of
the covariates, around an intercept $\mu$:
$\operatorname{logit} P(Y_{ij} = 1) = \mu + \alpha_i + Z_{ij}^\top \beta$ and
$Y_{ij} = \mu + \alpha_i + Z_{ij}^\top \beta + \epsilon_{ij}$. The model pools information
across providers and shrinks the estimated effects toward 0, the more so the larger the
residual variance is relative to $\sigma_\alpha^2$ and the fewer observations a provider
has.

**Correlated random effects.** Each covariate $x$ named in `within_between` is replaced by
its provider mean $\bar x_i$ and the deviation $x_{ij} - \bar x_i$, each with its own
coefficient (Mundlak 1978), so that the effects may be correlated with the provider means.
The means are computed over every row of the data with a value of $x$, before rows with
missing values are dropped.

**Data.** Rows with a missing outcome, provider, or covariate are dropped. The design
matrix is that of `model.matrix()` with the session's contrasts. Providers are ordered as
`factor()` orders their IDs (numbers numerically, strings in the collation order of the
session's locale). The logistic fixed-effect models exclude providers with fewer than
`min_provider_size` complete observations (10 by default); the other models keep every
provider.

## Estimation

### Logistic fixed effects

The log-likelihood is
$$
\ell(\gamma, \beta) = \sum_{i=1}^m \sum_{j=1}^{n_i}
\left\{ Y_{ij} (\gamma_i + Z_{ij}^\top \beta) - \log\left(1 + e^{\gamma_i + Z_{ij}^\top \beta}\right) \right\}.
$$
With $p_{ij} = \operatorname{plogis}(\gamma_i + Z_{ij}^\top \beta)$ and weights
$w_{ij} = p_{ij}(1 - p_{ij})$, the information matrix has a diagonal block for the provider
effects, $D = \operatorname{diag}(D_i)$ with $D_i = \sum_j w_{ij}$; a cross block with
columns $B_i = \sum_j w_{ij} Z_{ij}$; and the covariate block
$I_{\beta\beta} = \sum_{ij} w_{ij} Z_{ij} Z_{ij}^\top$. With thousands of providers, inverting
the whole matrix is the bottleneck of Newton's method. The serial blockwise inversion
Newton algorithm (SerBIN, `method = "serbin"`, Wu et al. 2022) inverts it through its
blocks: with the Schur complement $S = I_{\beta\beta} - \sum_i B_i B_i^\top / D_i$ and the
scores $U_{\gamma_i} = \sum_j (Y_{ij} - p_{ij})$ and $U_\beta = \sum_{ij} (Y_{ij} - p_{ij}) Z_{ij}$,
the Newton step is
$$
d_\beta = S^{-1} \Big( U_\beta - \sum_i B_i U_{\gamma_i} / D_i \Big), \qquad
d_{\gamma_i} = \big( U_{\gamma_i} - B_i^\top d_\beta \big) / D_i .
$$
The block ascent Newton algorithm (`method = "ban"`, He et al. 2013) alternates a Newton
step for the provider effects, with $\beta$ fixed, and one for $\beta$.

The code follows pprof 1.0.3 in these details:

- Every provider effect starts at the logit of the overall event rate, and $\beta$ at 0.
- Weights that are exactly 0 are replaced by $10^{-20}$ in the provider block and, for
  SerBIN, in the covariate block.
- With `backtrack = TRUE`, the step is shortened by Armijo backtracking: its length $v$
  starts at 1 and is multiplied by 0.6 until
  $\ell(\theta + v d) - \ell(\theta) \ge 0.01\, v\, U^\top d$.
- After every update, each effect is kept within `effect_bound` ($b = 10$) of the median of
  the updated effects: $\gamma_i \leftarrow \min\{\max(\gamma_i, \tilde\gamma - b), \tilde\gamma + b\}$,
  with $\tilde\gamma$ the median of the effects just computed. (pprof 1.0.3's vignette said
  the median of the previous iteration's effects; the code has always used the current one.)
- There are three convergence criteria: the largest absolute change of a coefficient, the
  change of the log-likelihood relative to its new value, $|\Delta\ell / \ell|$, and the
  change relative to the gain since the starting values, $|\Delta\ell / (\ell - \ell_0)|$.
  The stopping rule `"any"` (the default) stops as soon as the smallest of the three is below
  `tol`, `"all"` when the largest is, and the other rules use one criterion. Because the two
  log-likelihood criteria shrink as the data grow, a fit on large data with `"any"` can stop
  while the provider effects still change. SerBIN runs at most `max_iter` + 1 iterations,
  BAN at most `max_iter`; a fit that reaches the limit warns.

A provider without events has an effect whose maximum likelihood estimate is
$-\infty$ (one with only events, $+\infty$). The bound keeps such effects finite, but each
iteration moves them further toward it, so their estimates depend on when the fit stops,
and Wald tests and intervals are unreliable for them.

The variances use probabilities kept within $[10^{-10}, 1 - 10^{-10}]$:
$\operatorname{Var}(\hat\beta) = S^{-1}$ and
$\operatorname{Var}(\hat\gamma_i) = 1 / D_i + J_i^\top S^{-1} J_i$ with $J_i = B_i / D_i$.
The log-likelihood counts $m + p$ parameters in AIC and BIC, and the AUC is that of the
fitted probabilities.

### Firth's penalty

`fit_logistic_firth()` maximizes the penalized log-likelihood
$\ell^*(\gamma, \beta) = \ell(\gamma, \beta) + \tfrac12 \log \det I(\gamma, \beta)$ (Firth 1993),
whose score adds to each residual $Y_{ij} - p_{ij}$ the term $h_{ij}(1/2 - p_{ij})$, with
$h_{ij}$ the leverage of observation $ij$. Each iteration takes a full Newton step without
a line search and keeps the effects within the bound. The fit stops when the largest change
of a coefficient is below `tol`, the only stopping rule, or after `max_iter` iterations
(1,000 by default). Weights that are exactly 0 are replaced by $10^{-10}$. As in pprof 1.0.3, the
variances, the log-likelihood, AIC, and BIC are those of the unpenalized likelihood at the
Firth estimates; the penalized log-likelihood is kept as `penalized_loglik`.

The penalty removes the leading term of the bias of the maximum likelihood estimates
(Firth 1993) and gives finite estimates when a provider's outcomes are all the same
(Heinze and Schemper 2002); `vignette("models", package = "pprof")` shows both fits for
the example's providers without events.

### Linear fixed effects

To avoid a design matrix with a column for every provider, `fit_linear_fe()` profiles the
provider effects out. For fixed $\beta$, the least-squares estimate of $\gamma_i$ is
$\bar Y_i - \bar Z_i^\top \beta$, and the coefficients come from the data centered within
providers, $\tilde Y_{ij} = Y_{ij} - \bar Y_i$ and $\tilde Z_{ij} = Z_{ij} - \bar Z_i$:
$$
\hat\beta = \Big( \sum_{ij} \tilde Z_{ij} \tilde Z_{ij}^\top \Big)^{-1} \sum_{ij} \tilde Z_{ij} \tilde Y_{ij},
\qquad
\hat\gamma_i = \bar Y_i - \bar Z_i^\top \hat\beta .
$$
Under the model these estimates are unbiased. The residual variance is
$\hat\sigma^2 = \mathrm{SSR} / (n - m - p)$, and
$\operatorname{Var}(\hat\beta) = \hat\sigma^2 (\sum_{ij} \tilde Z_{ij} \tilde Z_{ij}^\top)^{-1}$.
The variance of $\hat\gamma_i$ is
$\hat\sigma^2 \{ 1 / n_i + \bar Z_i^\top (\sum_{ij} \tilde Z_{ij} \tilde Z_{ij}^\top)^{-1} \bar Z_i \}$;
`provider_variance = "full"` uses this, and `"simplified"`, the default, its first term
$\hat\sigma^2 / n_i$, which leaves out the uncertainty of $\hat\beta$. The log-likelihood is
that of the normal model at $\hat\sigma^2 = \mathrm{SSR} / n$, with $m + p + 1$ parameters in
AIC and BIC.

### Random and correlated random effects

The random-effect models are fitted by lme4 (Bates et al. 2015): `lmer()` (restricted
maximum likelihood) and `glmer()` with the binomial family (Laplace approximation), with
the formula `outcome ~ covariates + (1 | provider)`. The provider effects are lme4's
conditional modes $\hat\alpha_i$. Their standard deviations are the conditional ones: for
the linear random-effect model, $s_i = \sqrt{R_i \hat\sigma^2 / n_i}$ with the shrinkage
factor $R_i = \hat\sigma_\alpha^2 / (\hat\sigma_\alpha^2 + \hat\sigma^2 / n_i)$, and for the
other models those of lme4.

## Provider tests

The null value, or norm, is the median of the estimated effects for fixed-effect models
(`null = "median"`, the default), for linear fixed-effect models also their mean weighted
by provider size (`"mean"`), and for random-effect models a number, 0 by default. A
two-sided test computes the upper-tail probability $P$ of the observed statistic under the
null, reports the p-value $2 \min(P, 1 - P)$, and flags the provider 1 (higher than
expected) if $P < a/2$, -1 (lower) if $P > 1 - a/2$, and 0 otherwise. A one-sided test
reports the probability of the tested tail and flags the provider in that direction when
it is below $a$.

For logistic fixed-effect models, the null probability of observation $ij$ is
$p^0_{ij} = \operatorname{plogis}(\gamma_0 + Z_{ij}^\top \hat\beta)$, kept within
$[10^{-10}, 1 - 10^{-10}]$ by the exact, bootstrap, and modified score tests, and
$O_i = \sum_j Y_{ij}$ is the provider's number of events.

- **Exact** (the default). Under the null, $O_i$ is a sum of independent Bernoulli
  variables with probabilities $p^0_{ij}$, so it has a Poisson-binomial distribution, and
  the test asks how likely a count as extreme as the observed $o$ is. The two-sided test
  uses the mid-p probability $P = P(O_i > o) + \tfrac12 P(O_i = o)$, which counts half the
  probability of the observed count and so is less conservative than counting all of it;
  the statistic is $z_{1-P}$. The one-sided tests use $P(O_i \ge o)$ (`"greater"`) and
  $P(O_i \le o)$ (`"less"`).
- **Bootstrap.** The same distribution, simulated: `n_resamples` totals per provider,
  each a sum of Bernoulli draws with probabilities $p^0_{ij}$, give $P$ as the share of
  totals above $o$ plus half the share equal to it (one-sided: at least $o$, at most
  $o$). The test estimates the exact test's probabilities by simulation, with a Monte Carlo
  error that shrinks as `n_resamples` grows (pprof 1.0.3's vignette called the two
  consistent). The draws use R's random number generator, so set a seed.
- **Score.** The modified score test (`score_type = "modified"`, the default) uses
  $z_i = \sum_j (Y_{ij} - p^0_{ij}) / \sqrt{\sum_j p^0_{ij}(1 - p^0_{ij})}$ with the
  coefficients $\hat\beta$ of the full model, instead of refitting the model under the null
  for each provider. The test called standard (`score_type = "standard"`) does not refit
  either: at the full model's estimates, it divides the same score, computed with
  probabilities that are not kept within the bounds, by $\sqrt{V_i}$ with
  $V_i = I_{ii} - I_{\beta i}^\top S_i^{*-1} I_{\beta i}$, where
  $I_{ii} = \sum_j p^0_{ij}(1 - p^0_{ij})$, $I_{\beta i} = \sum_j p^0_{ij}(1 - p^0_{ij}) Z_{ij}$,
  and $S_i^*$ is the Schur complement of the full model's information with the null weights
  in place of provider $i$'s weights and without provider $i$'s effect. The variance so
  accounts for the estimation of $\beta$ and of the other providers' effects.
- **Wald.** $z_i = (\hat\gamma_i - \gamma_0) / \operatorname{se}(\hat\gamma_i)$, compared with
  the normal distribution. It is not valid for providers without events or with only
  events, whose estimates depend on when the fit stopped, and the function warns when it
  covers them.

The two-sided p-value of the exact test, computed by hand with the poibin package for the
first hospital, and the bootstrap test against the exact test:

```{r exact-by-hand}
tests <- test_providers(fit)
rows <- provider_index(fit) == 1
p0 <- plogis(tests$null_value + linear_predictor(fit)[rows])
p0 <- pmin(pmax(p0, 1e-10), 1 - 1e-10)
o <- sum(observed_outcome(fit)[rows])
upper <- 1 - poibin::ppoibin(o, p0) + 0.5 * poibin::dpoibin(o, p0)
c(by_hand = 2 * min(upper, 1 - upper), pprof = tests$table$p_value[1])
set.seed(1)
bootstrap <- test_providers(fit, test = "bootstrap", n_resamples = 2000)
max(abs(bootstrap$table$p_value - tests$table$p_value))
```

With 2,000 resamples, the bootstrap p-values differ from the exact ones by up to a few
hundredths, the order of their Monte Carlo error.

**Linear fixed effects.** $z_i = (\hat\gamma_i - \gamma_0) / \operatorname{se}(\hat\gamma_i)$,
compared with the normal distribution with the simplified variance and with Student's t
with $n - m - p$ degrees of freedom with the full variance, as in pprof 1.0.3.

**Random and correlated random effects.**
$z_i = (\hat\alpha_i - \alpha_0) / s_i$, compared with the normal distribution, with the
conditional standard deviation $s_i$. When lme4 estimates $\sigma_\alpha^2$ as 0, every
$\hat\alpha_i$ and $s_i$ is 0 and every flag is missing.

### What the random-effect tests and intervals do

The random-effect statistic divides a shrunk estimate by its conditional standard
deviation, and this sets how often it flags a provider. In the linear random-effect model,
with the variance components taken as known, $\hat\alpha_i = R_i (\alpha_i + \bar\epsilon_i)$,
where $\bar\epsilon_i$ is the provider's mean error, with variance $\sigma^2 / n_i$. For a
provider without an effect ($\alpha_i = 0$), $\hat\alpha_i / s_i$ is normal with mean 0 and
variance $R_i < 1$, so the test flags it with probability
$2\{1 - \Phi(z_{1-a/2} / \sqrt{R_i})\}$, less than $a$, and the less the smaller the
provider. The interval $\hat\alpha_i \pm z_{1-a/2}\, s_i$ is the posterior interval of
$\alpha_i$: averaged over effects drawn from $N(0, \sigma_\alpha^2)$, it covers the effect
with probability $1 - a$, but for a given provider the coverage depends on the effect,
above $1 - a$ for an effect near 0 and below it, down to 0, for a large one, because the
estimate is shrunk toward 0 (Morris 1983).

For the linear random-effect fit of `ExampleDataLinear`, whose providers have 54 to 99
observations, and for the same data cut to 10 observations per provider:

```{r re-size, message = FALSE}
data(ExampleDataLinear)
continuous <- data.frame(y = ExampleDataLinear$Y, hospital = ExampleDataLinear$ProvID, ExampleDataLinear$Z)
flag_probability <- function(model) {
  variance <- model$variance_components$provider
  shrinkage <- variance / (variance + model$sigma^2 / provider_table(model)$n_obs)
  probability <- 2 * (1 - pnorm(qnorm(0.975) / sqrt(shrinkage)))
  c(shrinkage_min = min(shrinkage), shrinkage_max = max(shrinkage),
    flag_probability_min = min(probability), flag_probability_max = max(probability))
}
re_fit <- fit_linear_re(y ~ z1 + z2 + z3 + z4 + z5, continuous, "hospital")
flag_probability(re_fit)
first_ten <- continuous[ave(seq_len(nrow(continuous)), continuous$hospital, FUN = seq_along) <= 10, ]
flag_probability(fit_linear_re(y ~ z1 + z2 + z3 + z4 + z5, first_ten, "hospital"))
```

At level 0.95, a provider without an effect is flagged with probability
`r sprintf("%.1f%%", 100 * min(flag_probability(re_fit)[3:4]))` to
`r sprintf("%.1f%%", 100 * max(flag_probability(re_fit)[3:4]))` in the full data, and
about 1% with 10 observations per provider, against 5% for the fixed-effect test. The
logistic random-effect and the correlated random-effect models behave in the same way, the
logistic ones approximately.

A simulation study in the package's source repository (`validation/simulation-report.md`:
50 providers, 200 replicates per model, six outlying providers with effects of ±0.8 on the
log-odds scale or ±0.5 units, the other providers without an effect, each model at its
defaults) found:

| Models | Providers without an effect flagged | Outlying providers flagged | Intervals covering the effect: providers without an effect | Intervals covering the effect: outlying providers |
|---|---|---|---|---|
| fixed effects (exact or Wald test) | 5.0% to 5.5% | 77.2% to 86.4% | 94.5% to 94.8% | 95.0% to 96.1% |
| random and correlated random effects (Wald test) | 0.7% to 0.9% | 46.2% to 60.3% | 99.1% to 99.3% | 38.1% to 44.5% |

Without any provider effect, about half of the random-effect fits were singular, with every
flag missing.

The random-effect flags and intervals so answer a different question from the fixed-effect
ones, and the two kinds of flags are not interchangeable. Jones and Spiegelhalter (2011)
and Ohlssen, Sharples and Spiegelhalter (2007) discuss ways of identifying unusual
providers with hierarchical models.

### Many providers

Each provider is tested at the chosen level, without an adjustment for multiplicity, so
with $m$ providers about $a\, m$ are flagged by chance when no provider differs from the
norm (with calibrated tests). Jones, Ohlssen and Spiegelhalter (2008) discuss false
discovery rates for comparing many providers.

## Standardized measures

**Logistic fixed effects.** The indirectly standardized ratio of provider $i$ is
$O_i / E_i$, with the expected count
$E_i = \sum_j \operatorname{plogis}(\gamma_0 + Z_{ij}^\top \hat\beta)$: the number of events
expected for the provider's patients if the provider performed at the norm (He and
Schaubel 2014; Jones, Ohlssen and Spiegelhalter 2008). A ratio above 1 means more events
than expected. The directly standardized ratio of provider $k$ is $E^{(k)} / O$, with
$O = \sum_{ij} Y_{ij}$ and
$E^{(k)} = \sum_{ij} \operatorname{plogis}(\hat\gamma_k + Z_{ij}^\top \hat\beta)$, summed over
all observations: the number of events expected in the whole population if every provider
performed as provider $k$ (He 2019). It lies between 0 and $n / O$, the limits as
$\hat\gamma_k \to -\infty$ and $+\infty$. An indirect ratio compares a provider with the norm
for its own patients, so the expectations of providers with different patients differ;
direct ratios apply every provider's effect to the same patients. Rates are ratios times
the population rate $100\, O / n$, in percent, kept within 0 and 100 (a rate computed this
way can exceed 100%).

The indirect ratios, computed by hand:

```{r ratio-by-hand}
measures <- standardize_providers(fit, measure = "ratio")
expected <- plogis(measures$null_value + linear_predictor(fit))
by_hand <- tapply(observed_outcome(fit), provider_index(fit), sum) / tapply(expected, provider_index(fit), sum)
all.equal(as.vector(by_hand), measures$table$estimate)
```

**Linear fixed effects.** The indirect difference is $(O_i - E_i) / n_i$ with
$O_i = \sum_j Y_{ij}$ and $E_i = \sum_j (\gamma_0 + Z_{ij}^\top \hat\beta)$, and the direct
difference of provider $k$ is
$\{\sum_{ij} (\hat\gamma_k + Z_{ij}^\top \hat\beta) - \sum_{ij} (\gamma_0 + Z_{ij}^\top \hat\beta)\} / n$;
both equal the provider's $\hat\gamma - \gamma_0$ up to rounding.

**Random and correlated random effects.** The numerator of the indirect measures is the
sum of the model's fitted values over the provider's observations, which include the
estimated effect $\hat\alpha_i$, not the observed outcomes (predicted over expected, as in
pprof 1.0.3); the expected values use the null effect, 0 by default. Logistic models give
$\sum_j \hat p_{ij} / \sum_j \operatorname{plogis}(\hat\mu + \alpha_0 + Z_{ij}^\top \hat\beta)$
and the direct ratio $\sum_{ij} \operatorname{plogis}(\hat\mu + \hat\alpha_k + Z_{ij}^\top \hat\beta) / O$;
linear models the differences
$\{\sum_j \hat Y_{ij} - \sum_j (\hat\mu + \alpha_0 + Z_{ij}^\top \hat\beta)\} / n_i$ and
$\{\sum_{ij} (\hat\mu + \hat\alpha_k + Z_{ij}^\top \hat\beta) - \sum_{ij} Y_{ij}\} / n$.

## Intervals

**Provider effects of logistic fixed-effect models.** The exact and score intervals
invert the two-sided tests: the limits are the effects at which the test's $P$ equals
$a/2$ and $1 - a/2$, found with `uniroot()` in brackets that widen from the estimate in
steps of 5 (three attempts; a limit without a root is infinite). For a provider without
events, the interval is $(-\infty, U]$, and for one with only events $[L, +\infty)$; the
finite limit solves its equation at $a$, not $a/2$, as in pprof 1.0.3. The Wald interval is
$\hat\gamma_i \pm z_{1-a/2} \operatorname{se}(\hat\gamma_i)$.

**Standardized measures.** The limits of the effect are carried through the measure's
sums: for the indirect ratio,
$\big[\sum_j \operatorname{plogis}(\gamma_L + Z_{ij}^\top \hat\beta) / E_i,\;
\sum_j \operatorname{plogis}(\gamma_U + Z_{ij}^\top \hat\beta) / E_i\big]$, and for the
direct ratio the same sums over all observations, divided by $O$. A provider without events
gets the lower limit 0, and one with only events the upper limit $n_i / E_i$ (direct:
$n / O$). Because the expected count is that of the null effect, the interval of the
indirect ratio contains 1 exactly when the effect's interval contains $\gamma_0$, and the
interval of the direct ratio contains the direct ratio at $\gamma_0$ exactly when the
effect's interval contains $\gamma_0$. (pprof 1.0.3's vignette named the provider's own
direct ratio here.) One-sided
intervals use $z_{1-a}$ and leave the other side open ($\pm\infty$ for effects; 0 or the
upper bound for ratios).

**Linear fixed effects.** $\hat\gamma_i \pm c\, \operatorname{se}(\hat\gamma_i)$, with $c$ the
$1 - a/2$ quantile of Student's t with $n - m - p$ degrees of freedom with the simplified
variance and of the normal distribution with the full variance, the reverse of the tests,
as in pprof 1.0.3; the differences' intervals shift by the same amounts.

**Random and correlated random effects.** $\hat\alpha_i \pm z_{1-a/2}\, s_i$, carried
through the measures' sums.

The Wald interval of the effects, by hand (the function warns, because three hospitals have
no events):

```{r wald-by-hand}
effects <- provider_effects(fit, interval = "wald")
by_hand <- effects$table$estimate + qnorm(1 - (1 - 0.95) / 2) * effects$table$std_error
all.equal(by_hand, effects$table$upper)
```

## Covariates

Every model has Wald tests of the coefficients. Logistic fixed-effect models use
$z = (\hat\beta - \beta_0) / \operatorname{se}$, the p-value $2\{1 - \Phi(|z|)\}$, and the
interval $\hat\beta \pm z_{1-a/2} \operatorname{se}$; linear fixed-effect models Student's t
with $n - m - p$ degrees of freedom; linear random-effect and correlated random-effect
models Student's t with $n - p - m + 1$ degrees of freedom ($p$ counting the intercept) and
lme4's Wald intervals; logistic random-effect and correlated random-effect models the
p-value $2\{1 - \Phi(z)\}$, as in pprof 1.0.3, which exceeds 1 for negative estimates, and
lme4's Wald intervals.

For logistic fixed-effect models, `test_coefficients()` also gives likelihood-ratio and
score tests. As in pprof 1.0.3, the model without the tested covariate is refit with the
default settings of `fit_logistic_fe()`, whatever settings the model used; the
likelihood-ratio statistic is the difference of the two $-2\ell$, computed with the effects
kept within 10 of their median, and the score statistic is the squared score of the
covariate over its efficient information; both are compared with the chi-square
distribution with one degree of freedom. With one covariate, the model without it has
provider effects only, and the likelihood-ratio test is computed against it, as pprof 1.0.3
computes it from R 4.5.0; the score test is not available there, as pprof 1.0.3 fails on it.
Like pprof 1.0.3's, that fit stops after one iteration under the default stopping rule,
whose criterion for the change in the coefficients is 0 when there are none, so the
statistic is computed against provider effects one step from their starting values rather
than their maximum-likelihood estimates; the methodology owners decide whether to change
that.

## Funnel limits

For logistic fixed-effect models, the precision of provider $i$ is $E_i^2 / V_i$, with
$V_i = \sum_j p^0_{ij}(1 - p^0_{ij})$ (probabilities not kept within bounds), and the limits
at precision $w$ are $1 \pm z_{1-a/2} / \sqrt{w}$, the lower one at least 0. These are the
limits of the score statistic $(O_i - E_i) / \sqrt{V_i}$, and the providers are coloured by
the modified score test at the first level, so a provider is flagged when its point lies
outside them (Spiegelhalter 2005; Wu et al. 2023). For linear fixed-effect models, the
precision is $n_i$ and the limits are $0 \pm z_{1-a/2}\, \hat\sigma / \sqrt{n_i}$, with flags
from the Wald test; with the full provider variance, the limits still use
$\hat\sigma^2 / n_i$ and the normal distribution while the test uses the full variance and
Student's t, so a point outside the limits may not be flagged, as in pprof 1.0.3. Exact
funnel limits are not available, and random-effect models have no funnel plot.

## Numerical conventions

| Convention | Value |
|---|---|
| Weights that are exactly 0 (SerBIN, BAN, the standard score test) | replaced by $10^{-20}$ |
| Weights that are exactly 0 (Firth) | replaced by $10^{-10}$ |
| Probabilities in variances and in the exact, bootstrap, and modified score tests | kept within $[10^{-10}, 1 - 10^{-10}]$ |
| Armijo backtracking | sufficient increase 0.01, step multiplied by 0.6 |
| Provider effects | kept within `effect_bound` (10) of their median |
| Firth's log-determinant | provider blocks at least $10^{-12}$; ridge $10^{-8}$ if the Cholesky factorization fails |
| Interval search | brackets in steps of 5, three attempts |
| Rates | kept within 0 and 100 |

## References

Bates D, Mächler M, Bolker B, Walker S (2015). Fitting linear mixed-effects models using
lme4. *Journal of Statistical Software*, 67(1), 1-48. <https://doi.org/10.18637/jss.v067.i01>

Firth D (1993). Bias reduction of maximum likelihood estimates. *Biometrika*, 80(1),
27-38. <https://doi.org/10.1093/biomet/80.1.27>

He K (2019). Indirect and direct standardization for evaluating transplant centers.
*Journal of Hospital Administration*, 8(1), 9-14. <https://doi.org/10.5430/jha.v8n1p9>

He K, Kalbfleisch JD, Li Y, Li Y (2013). Evaluating hospital readmission rates in dialysis
facilities; adjusting for hospital effects. *Lifetime Data Analysis*, 19, 490-512.
<https://doi.org/10.1007/s10985-013-9264-6>

He K, Schaubel DE (2014). Semiparametric methods for center effect measures based on the
ratio of survival functions. *Lifetime Data Analysis*, 20(4), 619-644.
<https://doi.org/10.1007/s10985-014-9293-9>

Heinze G, Schemper M (2002). A solution to the problem of separation in logistic
regression. *Statistics in Medicine*, 21(16), 2409-2419. <https://doi.org/10.1002/sim.1047>

Jones HE, Ohlssen DI, Spiegelhalter DJ (2008). Use of the false discovery rate when
comparing multiple health care providers. *Journal of Clinical Epidemiology*, 61(3),
232-240. <https://doi.org/10.1016/j.jclinepi.2007.04.017>

Jones HE, Spiegelhalter DJ (2011). The identification of "unusual" health-care providers
from a hierarchical model. *The American Statistician*, 65(3), 154-163.
<https://doi.org/10.1198/tast.2011.10190>

Morris CN (1983). Parametric empirical Bayes inference: theory and applications. *Journal
of the American Statistical Association*, 78(381), 47-55.
<https://doi.org/10.1080/01621459.1983.10477920>

Mundlak Y (1978). On the pooling of time series and cross section data. *Econometrica*,
46(1), 69-85. <https://doi.org/10.2307/1913646>

Ohlssen DI, Sharples LD, Spiegelhalter DJ (2007). A hierarchical modelling framework for
identifying unusual performance in health care providers. *Journal of the Royal
Statistical Society: Series A*, 170(4), 865-890. <https://doi.org/10.1111/j.1467-985X.2007.00487.x>

Spiegelhalter DJ (2005). Funnel plots for comparing institutional performance.
*Statistics in Medicine*, 24(8), 1185-1202. <https://doi.org/10.1002/sim.1970>

Wu W, Kuriakose JP, Weng W, Burney RE, He K (2023). Test-specific funnel plots for
healthcare provider profiling leveraging individual- and summary-level information.
*Health Services and Outcomes Research Methodology*, 23(1), 45-58.
<https://doi.org/10.1007/s10742-022-00285-9>

Wu W, Yang Y, Kang J, He K (2022). Improving large-scale estimation and inference for
profiling health care providers. *Statistics in Medicine*, 41(15), 2840-2853.
<https://doi.org/10.1002/sim.9387>

Xia L, He K, Li Y, Kalbfleisch JD (2022). Accounting for total variation and robustness in
profiling health care providers. *Biostatistics*, 23(1), 257-273.
<https://doi.org/10.1093/biostatistics/kxaa024>
